### 2017-03-21 preproposal/preproposal.tex
added objectives and tentative schedule

Cancer cells injected in mice grow into tumors composed of distinct sub-populations of cells, which exhibit patterned levels of gene expression. In order to characterize sub-populations of tumor cells and identify useful gene modules, we propose applying Bayesian nonparametrics to model single-cell RNA-sequencing (scRNA-seq) data from mouse tumors. Parametric topic models based on matrix factorization have shown promise in identifying meaningful clusters of cells and genes in heterogeneous cell populations. However, the need to pre-determine the number of clusters is a major limitation, particularly when extending these models to time-series data, as sub-populations of cells may appear, disappear, or diverge in expression over time. We propose exploring the utility of nonparametric models based on the Hierarchical Dirichlet Process (HDP) and the Indian Buffet Process (IDP) for modeling scRNA-seq data from single time points. We plan to compare their performance to that of existing methods for scRNA-seq analysis, as well as that of the top parametric topic model, Latent Dirichlet Allocation (LDA). Time permitting, we would also like to explore the extension of these models to time-series data, in order to understand how sub-populations of cells in the tumor microenvironment evolve over time.

>>>>

Cancer cells injected in mice grow into tumors composed of distinct sub-populations of cells, which exhibit patterned levels of gene expression. In order to characterize sub-populations of tumor cells and identify useful gene modules, we propose applying Bayesian nonparametrics to model single-cell RNA-sequencing (scRNA-seq) data from mouse tumors. Parametric topic models based on matrix factorization have shown promise in identifying meaningful clusters of cells and genes in heterogeneous cell populations; however, the need to pre-determine the number of clusters is a major limitation, particularly when extending these models to time-series data, as sub-populations of cells may appear, disappear, or diverge in expression over time. We propose exploring the utility of nonparametric models based on the Hierarchical Dirichlet Process (HDP) and the Indian Buffet Process (IDP) for modeling scRNA-seq data from single time points. We plan to compare their performance to that of existing methods for scRNA-seq analysis, as well as that of the top parametric topic model, Latent Dirichlet Allocation (LDA). Time permitting, we would also like to explore the extension of these models to time-series data, in order to understand how sub-populations of cells in the tumor microenvironment evolve over time.

### 2017-03-21 proposal/proposal.tex
add preproposal
added objectives and tentative schedule

\begin{centering}
\Large
    \textbf{Time-series and Hierarchical Modeling for sc-RNA Expression Levels}  \\
    \vspace{2mm}
    \normalsize
    Skanda Koppula (\url{skoppula@mit.edu}), Karren Yang (\url{karren@mit.edu}) \\
    \vspace{2mm}
    \normalsize
    6.882 Project Pre-Proposal, \today \\
\end{centering}
\vspace{5mm}

Cancer cells implanted into mice exhibit patterned levels of RNA expression. The expression in each cancer cell varies over time, and the cells may naturally diverge in their expression during this time. Previous work has looked into an LDA-based clustering of expression vectors for particular time slices, resulting in interesting and biologically meaningful cell grouping shifts over time. \\

This project is interested in constructing the first generative model of this process. Specifically, we look to answer some of the following questions: 
\begin{enumerate}
\item Can we apply a topic model based probabalistic model for a collection of expression vectors for a particular time slice? Will a topic model based solution yield meaningful cell type assignments?
\item How can we apply time-series probabilistic models such as Chinese Restaurant to model cell types over time?
\item How can we model noise in expression level observations in our probabilistic models? Is it possible to use our model to establish estimates for missing coordinates in expression vectors (a common problem in high-throughput expression measurement machines)?
\item Is it possible to emulate expression data with a simple sampling processes (for the purposes of testing any models we create)?
\end{enumerate}
These were a few lines of inquiry we were interested in pursuing. The authors have access to an sc-RNA expression level dataset through their research lab.
\end{document}

>>>>

\begin{centering}
\Large
    \textbf{Bayesian Clustering and Topic Modeling for Analyzing Single-Cell RNA Sequencing Data}  \\
    \vspace{2mm}
    \normalsize
    Skanda Koppula (\url{skoppula@mit.edu}), Karren Yang (\url{karren@mit.edu}) \\
    \vspace{2mm}
    \normalsize
    6.882 Project Proposal, \today \\
\end{centering}
\vspace{5mm}

\begin{abstract}

Tumors are composed of many sub-populations of cells, which exhibit different patterns of gene expression. In order to characterize sub-populations of tumor cells and identify functional gene modules, we propose applying Bayesian methods (parametric and non-parametric) for clustering and topic modeling to single-cell RNA-sequencing (scRNA-seq) data from tumors. We will compare these methods to standard methods for analyzing scRNA-seq data based their ability to identify known cell sub-populations and gene modules from the literature as well as their robustness to noise. We will furthermore determine if integrating the clustering model with the topic model results in superior performance in these tasks. Time permitting, we will attempt to extend these models to accommodate time-series data.

\end{abstract}

\section{Overview}

Using single-cell RNA-sequencing (scRNA-seq) data, it is possible to obtain the expression levels of 10,000+ genes for 1000+ cells in a single experiment. Given such data, we are primarily interested in answering the following questions:

\begin{enumerate}
    \item Can Bayesian nonparametric topic models, such as the Hierarchical Dirichlet Process (HDP) and the Indian Buffet Process Compound Dirichlet Process (IBP-CDP), extract functional gene modules from this data? Do Bayesian nonparametric topic models perform as well as parametric topic models, such as Latent Dirichlet Allocation (LDA), on this data?
    \item Can we get meaningful cell-type assignments by clustering cells, using finite or infinite mixture models, based on the output of the topic models?
    \item How robust are these methods to dropout, a phenomenon commonly observed in real scRNA-seq data?
    \item Can integrated models of cell clustering and topic modeling yield superior results in these tasks?
    \item Can we extend any of these models to time-series data?

\end{enumerate}

Risk: For a 6-7 week project, this list seems ambitious, but we have mitigated the risk in several ways. (1) Although we are primarily interested in the utility of nonparametric methods for topic (gene module) modeling, we are also using Latent Dirichlet Allocation so that subsequent steps can proceed even if the nonparametric methods do not work well. (2) If we become stuck on steps 4-5, we can submit a good report from objectives 1-3.

\section{Tentative Schedule}
\subsection{Apply LDA and HDP to scRNA-seq data - SK}
\subsection{Apply IBP-CDP to scRNA-seq data - KY}
\subsection{Evaluate biological relevance of gene modules from topic models - KY}
\subsection{Evaluate robustness of topic models to dropout - SK}
\subsection{Apply finite mixture model to cluster topic model output - SK}
\subsection{Apply infinite mixture model to cluster topic model output - KY}
\subsection{Integrate topic modeling and cell clustering - TBA}
\subsection{Extend model to time-series data - TBA}

\begin{enumerate}
    \item Referenced paper, outside/related work
    \item Outline work to be done
    \item 4 steps per team member, internal deadlines
    \item Risks: what might be more difficult than planned, thoughts to mitigate these
    \item How to evaluate our methods?
\end{enumerate}
\end{document}
